---
title: "Prediction Assignment Writeup - Module 8: Practical Machine Learning"
author: "Roddy Mendoza Marriott"
date: "2026-02-09"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

**Background**

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

---

**The data**

-The training data for this project are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv]

-The test data are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv]

-The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har]. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

---

**Some insights about the project**

The goal of your project is to predict the manner in which they did the exercise. This is the `classe` variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.


---

**Data processing**

First, loading the packages and libraries.

```{r, message=FALSE,warning=FALSE}
library(plyr)
library(dplyr)
library(lattice)
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(kernlab)
library(randomForest)
library(knitr)
library(e1071)
```

---

**Getting and cleaning data**

```{r, message=FALSE,warning=FALSE}
trainingst <- read.csv("pml-training.csv")
testingst <- read.csv("pml-testing.csv")

```

```{r, message=FALSE,warning=FALSE}
dim(trainingst)
```


```{r, message=FALSE,warning=FALSE}
dim(testingst)
```


---

**Preprocessing and cleaning data**

we should Exclude the obvious columns i.e `X`, `user_name`,`raw_timestamp_part_1`,`raw_timestamp_part_2`, `cvtd_timestamp`,`roll_belt` which are the first 7 columns. We should also delete missing values and variables with near zero variance.

```{r, message=FALSE,warning=FALSE}
#Deleting missing values 
trainingst <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!",""))  
testingst <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!",""))
```



```{r, message=FALSE,warning=FALSE}
#Deleting missing values
trainingst<-trainingst[,colSums(is.na(trainingst)) == 0]
testingst <-testingst[,colSums(is.na(testingst)) == 0]
```


```{r, message=FALSE,warning=FALSE}
#Removing columns that are not predictors, which are the the seven first columns
trainingst   <-trainingst[,-c(1:7)]
testingst <-testingst[,-c(1:7)]
```

```{r, message=FALSE,warning=FALSE}
dim(trainingst)
```


```{r, message=FALSE,warning=FALSE}
dim(testingst)
```

From the above code block sum(completeCase) == nrows confirm that the number of complete case is equal to number of rows in `trainingdf` same for `testingdf`

Now we have only 53 columns(features) are left. we can preproccess the training and testing i.e converting into scales of `0` to `1` and replacing any `NA` values to average of that columns.


---

**Partition the data set into `training` and `testing` data from `trainingst`**

```{r, message=FALSE,warning=FALSE}
inTrain <- createDataPartition(y = trainingst$classe, p=0.75, list = FALSE)
training <- trainingst[inTrain, ]
testing <- trainingst[-inTrain, ]
```


---

**Training the model**

Two methods will be applied to model, and the best one will be used for the(testingst) predictions.

The methods are: `Decision Tree` and `Random Forests`.


**Model 1: Training the model with Decision Trees**

```{r, message=FALSE,warning=FALSE}
set.seed(40000)
fitDT <- rpart(classe ~ .,training, method="class")
# Normal plot
rpart.plot(fitDT)
```

```{r, message=FALSE,warning=FALSE}
#Use model to predict classe in validation testing set
predictionDT <- predict(fitDT, testing, type = "class")
```


```{r, message=FALSE,warning=FALSE}
#Estimate the errors of the prediction algorithm in the Decision Tree model
cmdt <-confusionMatrix(as.factor(testing$classe), predictionDT)
cmdt
```


```{r, message=FALSE,warning=FALSE}
# Accuracy plot
plot(cmdt$table, col = cmdt$byClass, 
main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmdt$overall['Accuracy'], 4)))
```


**Model 2: Training the model using Random Forest**

```{r, message=FALSE,warning=FALSE}
rfModel <- randomForest(as.factor(classe)~., data=training)
# Summary of the model
rfModel
```

```{r, message=FALSE,warning=FALSE}
# Plot the variable importance
varImpPlot(rfModel)
```


```{r, message=FALSE,warning=FALSE}
# Confusion matrix with testing
predTesting <- predict(rfModel, testing)
rfcfm  <- confusionMatrix(as.factor(testing$classe), predTesting)
rfcfm
```



```{r, message=FALSE,warning=FALSE}
plot(rfcfm$table, col = rfcfm$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(rfcfm$overall['Accuracy'], 4)))
```


**Remarks**

  -`Decision Tree Model` is the worst model running, it has the low mean and the highest standard deviation.

  -`Random Forest Model` it has the highest mean accuracy and lowest standard deviation.

Depending on how your model is to be used, the interpretation of the kappa statistic might vary One common interpretation is shown as follows:

• Poor agreement = Less than 0.20

• Fair agreement = 0.20 to 0.40

• Moderate agreement = 0.40 to 0.60

• Good agreement = 0.60 to 0.80

• Very good agreement = 0.80 to 1.00




